---
title: "Infrastructure Investment & Jobs Act Funding Allocation"
author: "Cliff Lee"
date: "2023-09-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(scales)
library(plotly)
library(ggplot2)
library(pixiedust)
library(kableExtra)
```

## Brief

For the 2020 presidential runoff, Biden campaigned on his 'Build Back Better' that broadly focused on updating the country's infrastructure that most needed updating. In 2021, Congress passed the Infrastructure and Investment and Jobs Act (IIJA) to fund public projects for transportation, citizen welfare, border security and communications.

However, did each state receive funding based on its population? Or did states who voted Democratic receive more funding?

The below visualizations show two conclusions:

 * Less populated states received more funding than more populated ones
 * States that voted for Biden did not receive extra funding (no political bias)

In short, states with lower populations (and presumably greater need) received extra funding whether they voted for the current administration or not.

```{r, echo=FALSE, message=FALSE}

## read the census population data
## territories from: https://www.census.gov/library/stories/2021/10/first-2020-census-united-states-island-areas-data-released-today.html
census <- read_csv('NST-EST2022-ALLDATA.csv')
census <- census %>% filter(STATE > 0)

census <- census %>% select(NAME, POPESTIMATE2022)
census$NAME <- census$NAME %>% str_to_upper()
colnames(census) <- c("state", "population")
#census %>% arrange(desc(population)) %>% print(n=100)

territories <- data.frame(
    state=c("AMERICAN SAMOA","GUAM","NORTHERN MARIANA ISLANDS","US VIRGIN ISLANDS"),
    population = c(49710 , 153836, 47329 ,87146)
)

census <- rbind(census, territories)

sum_population <- sum(census$population)

census <- census %>% mutate(estimate_allocation = population / sum_population)



# read Infrastructure Investment and Jobs excel data
iija <- read_excel("IIJA FUNDING AS OF MARCH 2023(1).xlsx")
colnames(iija) <- c("state", "allocation")
sum_allocation <- sum(iija$allocation)
iija <- iija %>% mutate (actual_allocation = allocation / sum_allocation)



total <- census %>% inner_join(iija)

# scraped from  https://www.archives.gov/electoral-college/2020
election <- read_csv("2020-election.csv")
election$state <- election$state %>% str_to_upper()
total <- total %>% inner_join(election)

total <- total %>% mutate (bias = estimate_allocation - actual_allocation)
#print(total, n=100)

#t %>% select(state, bias, biden, ratio_allocation, ratio_population) %>% print(n=1000)
total <- total %>% mutate (bias = ifelse(actual_allocation > estimate_allocation , 1, 0))
total <- total %>% mutate (biden = ifelse(biden > 0, 1, 0))
```


```{r, echo=FALSE, message=FALSE}

# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)
```


## Allocation Distribution

The chart below shows IIJA funding if based on state population versus the actual funding. We see more populated states received less funding then less populated ones.
```{r, echo=FALSE, message=FALSE}
t <- data.frame(population = total$population, Type="Population Based", allocation = total$estimate_allocation)
t2 <- data.frame(population = total$population, Type="Actual", allocation = total$actual_allocation)
t3 <- rbind (t,t2)

p <- ggplot(t3, aes(x=population,y=allocation, color=Type)) + geom_point() 
```

```{r, echo=FALSE, message=FALSE}
l <- lm(data=total, formula = actual_allocation ~ population)


p + geom_abline(slope=(total$estimate_allocation[1] / total$population[1]), intercept=0,color="seagreen3") +
geom_abline(slope=l$coefficients[2], intercept=l$coefficients[1], color="orangered2") +
  scale_y_continuous(labels = scales::percent, limits=c(0,0.12)) +
  scale_x_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  theme(panel.background = element_blank(), axis.line=element_line(color="grey")) +
  xlab("State Population") + ylab("Allocation of Total Budget") + ggtitle("IIJA State Funding") +
  theme(plot.title = element_text(hjust = 0.5))
```

## IIJA Agenda

The main goals of the IIJA are to "rebuild roads and bridges, improve public transit, replace lead pipes and address drinking water contamination, expand access to high-speed internet, and more." As such, we can see from figure 2 where the funding bias favors the middle of the US and territories.  

The map below shows how many states received funding by more than its estimated population-percentage (based on the year 2020). For example, Maine has a population of 1.3 million (the US had a total population of 332 million). If the IIJA funds were population based, then Maine would have received 0.417% of the total funds, however instead it received 0.561% (1.34% more). 

Mainly midwestern states received funds as well Alaska, Hawaii and a few coastal states. 


```{r, echo=FALSE, message=FALSE}
total$hover <- with(total, paste(state) )
## IIJA bias funding
iija_figure <- plot_geo(total, locationmode = 'USA-states') 
iija_figure <- iija_figure %>% add_trace(
    z = ~bias, locations = ~code, text = ~hover,
    color = ~bias, colors = 'Greens'
  )
iija_figure <- iija_figure %>% colorbar(title = "Funding Bias ")
iija_figure <- iija_figure %>% layout(
    title = 'States That Received Excess IIJA Funding',
    geo = g
  )
iija_figure
```


## Political Influence on Allocation

Another possible explanation for IIJA funding disbursement is the Biden administration favored states that he won during the 2020 election. The electoral map below shows the states that he won in 2020 and they are mainly on the coasts with a key states in the midwest. We can see the amount of overlap between this figure and the previous figure is minimal.

```{r, echo=FALSE, message=FALSE}
### Biden electoral votes
total$hover <- with(total, paste(state) )
election_figure <- plot_geo(total, locationmode = 'USA-states')
election_figure <- election_figure %>% add_trace(
    z = ~biden, locations = ~code, text = ~hover,
    color = ~biden, colors = 'Blues'
  )
election_figure <- election_figure %>% colorbar(title = "Biden Victory")
election_figure <- election_figure %>% layout(
    title = 'States Who Voted Democratic in 2020',
    geo = g
  )

election_figure
```

## States With Extra Funding

There are ten states that voted Democratic and received extra funding.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
states_biden <- election %>% filter(biden > 0) %>% select(state)
states_bias <- total %>% filter(actual_allocation > estimate_allocation) %>% select(state)
states_common <- states_bias %>% inner_join(states_biden)
bias_funding <- total %>% inner_join(states_common) %>% select (state, allocation, actual_allocation, estimate_allocation)

bias_funding <- bias_funding %>% 
  mutate(percent = round(x=((actual_allocation - estimate_allocation)*100),digits=2)) %>%
  arrange(desc(percent))

bias_funding <- bias_funding %>% select(-c('allocation','actual_allocation','estimate_allocation')) 

dust(bias_funding) %>%
  sprinkle_colnames(state = "Democratic State", percent = "Percent Extra Funding") %>% 
  kable() %>% 
  kable_styling()
```

For 14 states that voted against Biden, they received more funding than their population predicts.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
states_trump <- election %>% filter(trump > 0) %>% select(state)
states_bias <- total %>% filter(actual_allocation > estimate_allocation) %>% select(state)
states_common <- states_bias %>% inner_join(states_trump)
bias_funding <- total %>% inner_join(states_common) %>% select (state, allocation, actual_allocation, estimate_allocation)

bias_funding <- bias_funding %>% 
  mutate(percent = round(x=((actual_allocation - estimate_allocation)*100),digits=2)) %>%
  arrange(desc(percent))

bias_funding <- bias_funding %>% select(-c('allocation','actual_allocation','estimate_allocation')) 

dust(bias_funding) %>%
  sprinkle_colnames(state = "Republican State", percent = "Percent Extra Funding") %>% 
  kable() %>% 
  kable_styling()
```


### References

 * Election results - https://www.archives.gov/electoral-college/2020
 * Population measurements - https://www.census.gov/library/stories/2021/10/first-2020-census-united-states-island-areas-data-released-today.html